---
title: "Harmonize initial datasets for Alicia's thesis"
subtitle: "JCyL forest service data and SFNI data"
author: "Aitor Vázquez Veloso"
date: 2026-02-18
format: html
---

# Initial setup

```{r}
library(tidyverse)
library(openxlsx)
library(sf)
```

# Harmonize initial datasets

## Forest service data - forest management plans

Harmonize forest service data originally from forest management plans (*ordenaciones de montes*).

```{r}
# get df
trees_ord <- read.xlsx('../data/brutos/Datos_Fran/Ordenacion/Datos_Excel/Datos_forcipula.xlsx')
str(trees_ord)

# explore df
trees_ord <- trees_ord[!is.na(trees_ord$ALTURA), ]
summary(trees_ord$DIAMT_1)
summary(trees_ord$DIAMT_2)
summary(trees_ord$ALTURA)
table(trees_ord$ESPECIE)

# create codes
trees_ord$INVENTORY_ID <- 'JCyL_ORD'
trees_ord$PLOT_ID <- paste0(trees_ord$INVENTORY_ID, '_', trees_ord$PARCELA)
trees_ord$TREE_ID <- paste0(trees_ord$PLOT_ID, '_', trees_ord$N..ARBOL)

# manage variables
trees_ord$dbh <- (trees_ord$DIAMT_1 + trees_ord$DIAMT_2) / 20  # from mm to cm
trees_ord$h <- trees_ord$ALTURA/10  # from dm to m
trees_ord$species <- trees_ord$ESPECIE
trees_ord$expan <- 10000/(pi*trees_ord$RADIO^2)  # expansion factor

# species name
# https://github.com/aitorvv/SFNI_Spanish_Forest_National_Inventory-ready_to_use/blob/main/data/SFNI4_species_codes.csv
sfni_sp <- read.csv('../data/brutos/Datos_IFN/SFNI4_species_codes.csv')
trees_ord$species_name <- sfni_sp$Nombre_Especie[match(trees_ord$species, sfni_sp$Codigo)]
```

## Forest service data - tree selection

Harmonize forest service data originally from tree selection plans (*señalamientos de madera*).

```{r}
# get df
trees_sen <- read.csv('../data/brutos/Datos_Fran/Senhalamiento_madera/senhalamientos_madera_join.csv')
str(trees_sen)

# convert chr to num
trees_sen$dbh <- as.numeric(sub(",", ".", trees_sen$D..cm.))
trees_sen$h <- as.numeric(sub(",", ".", trees_sen$H..m.))

# explore df
summary(trees_sen$dbh)
summary(trees_sen$h)
table(trees_sen$SP)

# create codes
trees_sen$INVENTORY_ID <- 'JCyL_SEN'
trees_sen$PLOT_ID <- paste0(trees_sen$INVENTORY_ID, '_', trees_sen$ARCHIVO, '_', trees_sen$PARCELA, '_', trees_sen$RODAL,
                            '_', trees_sen$TRAMO)
trees_sen$TREE_ID <- paste0(trees_sen$PLOT_ID, '_', trees_sen$id_ARBOL)

# manage variables
trees_sen$expan <- NA  # not plots

# species name and code
sp_names <- c(
  Pn = "Pinus nigra",
  Ppr = "Pinus pinaster",
  Ps = "Pinus sylvestris",
  Qpy = "Quercus pyrenaica",
  Qsp = NA,
  Seco = NA
)
trees_sen$species_name <- unname(sp_names[trees_sen$SP])
trees_sen$species <- sfni_sp$Codigo[match(trees_sen$species_name, sfni_sp$Nombre_Especie)]
rm(sp_names)
```

## SFNI plots information

Filter SFNI plots to keep only those in the area of interest (central Palencia province and surrounding areas).

```{r}
# get df
# https://github.com/aitorvv/SFNI_Spanish_Forest_National_Inventory-ready_to_use/tree/main/data
load('../data/brutos/Datos_IFN/1.1_sfni_clean_data.rdata')
sfni_pal <- read.csv('../data/brutos/Datos_IFN/IFN_Palencia-target_plots.csv')  # SFNI plots in Palencia filtered on QGIS
sfni_le <- read.csv('../data/brutos/Datos_IFN/IFN_Leon-target_plots.csv')  # SFNI plots in León filtered on QGIS
plots_sfni <- rbind(sfni_pal, sfni_le)
# plots_sfni <- select(plots_sfni, -fid)
rm(sfni_pal, sfni_le, plots)

# filter sfni trees
trees_sfni <- trees[trees$PLOT_ID %in% plots_sfni$PLOT_ID, ]

# species name
trees_sfni$species_name <- sfni_sp$Nombre_Especie[match(trees_sfni$species, sfni_sp$Codigo)]
rm(trees, sfni_sp)
```

# Combine and complete tree datasets

Clean non-required columns, calculate missing variables, and combine all datasets into a single tree dataframe.

```{r}
# remove non-required columns
trees_ord <- select(trees_ord, INVENTORY_ID, PLOT_ID, TREE_ID, species, species_name, dbh, h, expan)
trees_sen <- select(trees_sen, INVENTORY_ID, PLOT_ID, TREE_ID, species, species_name, dbh, h, expan)
trees_jcyl <- rbind(trees_ord, trees_sen)
rm(trees_ord, trees_sen)

# get functions for tree variables:
# https://github.com/aitorvv/SFNI_Spanish_Forest_National_Inventory-ready_to_use/blob/main/scripts/0.0_support_tree_functions.r
source('../scripts/functions/0.0_support_tree_functions.r')

# get tree circumference (cm)
trees_jcyl$circumference <- get_circumference(trees_jcyl$dbh, dbh_units = 'cm')

# basal area (cm2) and basal area per hectare (m2/ha)
trees_jcyl$g <- get_g(trees_jcyl$dbh, dbh_units = 'cm')
trees_jcyl$g_ha <- get_g_ha(trees_jcyl$dbh, expan = trees_jcyl$expan, dbh_units = 'cm')

# slenderness
trees_jcyl$slenderness <- get_slenderness(trees_jcyl$dbh, trees_jcyl$h, dbh_units = 'cm', h_units = 'm')

# record dead (1) and alive trees (0)
trees_jcyl$dead <- 0

# remove functions
rm(get_mean_dbh, get_expan, get_circumference, get_g, get_g_ha, get_bal, get_slenderness, get_coord_rel, get_coord_utm,
   get_sfni_expan)

# check columns available at SFNI but not in JCyL datasets
setdiff(names(trees_sfni), names(trees_jcyl))

# adapt SFNI trees to JCyL datasets structure and skip non-required columns
trees_sfni$PLOT_ID <- trees_sfni$IFN_PLOT_ID
trees_sfni$TREE_ID <- trees_sfni$IFN_TREE_ID
trees_sfni <- select(trees_sfni, INVENTORY_ID, PLOT_ID, TREE_ID, species, species_name, dbh, h, expan, circumference, g, g_ha,
                     slenderness, dead, shape)

# combine datasets
trees_jcyl$shape <- NA  # add missing column to JCyL dataset
trees <- rbind(trees_jcyl, trees_sfni)
trees <- trees[!is.na(trees$species), ]
```

# Combine and complete plot datasets

Clean non-required columns, calculate missing variables, and combine all datasets into a single plot dataframe.

*Note: as JCyL 'señalamientos' datasets are not plots, only JCyL 'ordenaciones' and SFNI plots are used.*

```{r}
# get trees from JCyL 'ordenaciones' only
trees_ord <- trees_jcyl[trees_jcyl$INVENTORY_ID == 'JCyL_ORD', ]

# get functions for plot variables:
# https://github.com/aitorvv/SFNI_Spanish_Forest_National_Inventory-ready_to_use/blob/main/scripts/0.0_support_plot_functions.r
source('../scripts/functions/0.0_support_plot_functions.r')

# get N, DBH, G and G metrics for plots
plots_metrics <- get_plot_data(df_trees = trees_ord, plot_id_column_1 = 'PLOT_ID', plot_id_column_2 = NA, 
                               dbh_column = 'dbh', h_column = 'h', expan_column = 'expan', g_column = 'g')
# plots_metrics <- get_plot_data_basic(df_trees = trees_ord, plot_id_column_1 = 'IFN_PLOT_ID', plot_id_column_2 = NA, 
#                                      dbh_column = 'dbh', expan_column = 'expan', g_column = 'g')

# get dominant height (m) per plot
plots_Ho <- get_dominant_value(trees_ord, dominant_value_col = "h", output_col_name = "Ho",
                               plot_id_col = 'PLOT_ID', bigger_value_col = "h", expan_col = "expan")

# get dominant diameter (cm) per plot
plots_Do <- get_dominant_value(trees_ord, dominant_value_col = "dbh", output_col_name = "Do",
                               plot_id_col = 'PLOT_ID', bigger_value_col = "dbh", expan_col = "expan")

# get plot data by species
plots_sp <- get_plot_by_species(df_trees = trees_ord, plot_id_column_1 = 'PLOT_ID', 
                                species_column = 'species', expan_column = 'expan', g_column = 'g')

# get plot data according to tree status
plots_status <- get_plot_mortality(df_trees = trees_ord, plot_id_column_1 = 'PLOT_ID', 
                                   dead_column = 'dead', expan_column = 'expan', g_column = 'g')

# merge all the plot data previously calculated
plots_ord <- merge(plots_metrics, plots_Ho, by.x = 'PLOT_ID', by.y = 'PLOT_ID', all.x = TRUE)
plots_ord <- merge(plots_ord, plots_Do, by.x = 'PLOT_ID', by.y = 'PLOT_ID', all.x = TRUE)
plots_ord <- merge(plots_ord, plots_sp, by.x = 'PLOT_ID', by.y = 'PLOT_ID', all.x = TRUE)
plots_ord <- merge(plots_ord, plots_status, by.x = 'PLOT_ID', by.y = 'PLOT_ID', all.x = TRUE)

# remove temporal data
rm(plots_metrics, plots_Ho, plots_Do, plots_sp, plots_status)

# get plot slenderness
plots_ord$slenderness <- get_plot_slenderness(df_plots = plots_ord, hm_column = 'h_mean', dbhm_column = 'dbh_mean')

# get plot dominant slenderness
plots_ord$dominant_slenderness <- get_plot_dominant_slenderness(df_plots = plots_ord, Ho_column = 'Ho', Do_column = 'Do')

# get SDI
plots_ord$SDI <- get_SDI(df_plots = plots_ord, N_column = 'N', Dg_column = 'dg', r_value = 1.605)

# get Hart index for normal and sttagered plots
plots_ord$S <- get_hart_index(df_plots = plots_ord, Ho_column = 'Ho', N_column = 'N')
plots_ord$S_staggered <- get_hart_index_staggered(df_plots = plots_ord, Ho_column = 'Ho', N_column = 'N')

# remove functions
rm(get_plot_data, get_plot_by_species, get_plot_mortality, get_plot_slenderness, 
   get_plot_dominant_slenderness, get_SDI, get_hart_index, get_hart_index_staggered, get_wgs84_coordinates,
   get_Dg, get_100_bigger_trees, get_dominant_value, get_plot_data_basic)

# include missing code
plots_ord$INVENTORY_ID <- 'JCyL_ORD'
plots_ord <- select(plots_ord, INVENTORY_ID, everything())  # reorder columns

# check columns available at SFNI but not in JCyL datasets
setdiff(names(plots_sfni), names(plots_ord))

# adapt SFNI plots to JCyL datasets structure and skip non-required columns
plots_ord <- select(plots_ord, -hg, -species)

# correct SFNI plot code
plots_sfni$PLOT_ID <- plots_sfni$IFN_PLOT_ID
plots_sfni <- select(plots_sfni, names(plots_ord))

# combine datasets
plots <- rbind(plots_ord, plots_sfni)
rm(plots_sfni, plots_ord, trees_ord, trees_jcyl, trees_sfni)
```

# Save harmonized datasets

```{r}
save(trees, plots, file = '../data/procesados/1_harmonized_tree_plot_df_jcyl_sfni.rdata')
write.csv(trees, '../data/procesados/1_harmonized_tree_df_jcyl_sfni.csv', row.names = FALSE)
write.csv(plots, '../data/procesados/1_harmonized_plot_df_jcyl_sfni.csv', row.names = FALSE)
```
